---
title: "Knapsack experiments write-up"
author: "Molly Grove and Elliott Beach"
date: "March 5, 2016"
output: 
  html_document:
    toc: true
---

# Introduction

In this lab, we used three different population-based algorithms to find approximations for seven knapsack problems. 

For each iteration, our algorithm starts with a population of 100 and generates 300 children using one of three algorithms: mutation only, uniform crossover, and two-point crossover. The algorithm then combines all 400 answers (300 children and 100 parents) and chooses the best 100. 

The first algorithm for generating children uses only random mutation, with no crossover. The algorithm uses Nic's mutation method of starting with an answer and flipping each bit in :choices with probability of 1/n, where n is the length of the list. Flipping a bit adds or removes an item from the knapsack. 

The other two algorithms for generating children use crossover. We use tournament selection to choose the two parents, meaning that for each parent, we randomly choose two answers and use the one with the highest score as the parent. Unlike traditional crossover methods, our crossover methods generate only one child. For each crossover method, we tried regular crossover and combining crossover with mutating each child with a probability of 1/10. 

The second algorithm for generating children uses uniform crossover. It starts with two parent answers and loops through the :choices lists, randomly choosing which bit to use for the child answer. 

The third algorithm uses two-point crossover, which takes the first _m_ bits from the first parent's :choices list, the bits from _m_ to _n_ from the second parent, and the rest from the third parent to generate a child's :choices list. 

# Experimental setup

We used population sizes of 100 with 30 repetitions and 100,000 iterations on the following problems:

* `knapPI_11_20_1000_4`
* `knapPI_13_20_1000_4` 
* `knapPI_16_20_1000_4`
* `knapPI_11_200_1000_4`
* `knapPI_13_200_1000_4`
* `knapPI_16_200_1000_4`
* `knapPI_16_1000_1000_3`

(These names are abbreviated to, e.g., `11_20_1000_4`, in diagrams below.) We gathered data for random mutation, uniform crossover, uniform crossover with mutation, two-point crossover, and two point crossover with mutation. 

# Results

## A basic comparison of the search algorithms

The following is a plot comparing the scores of the search algorithms over all problems. 

```{r}
data_50_runs <- read.csv("../data/genetic/data", sep="")

par(mar=c(12,7,0,0))
plot(data_50_runs$Score ~ data_50_runs$Search_method,
     xlab="", ylab="Score", las=2)
```

The plot shows that uniform crossover was the best, while random mutation was the worst. However, the scores in the various problems were so different that it is hard to tell what is happening. 


```{r}
negs <- subset(data_50_runs, Score<0)
nrow(negs)
unique(negs$Problem)
```

So let's run a pairwise Wilcoxon test:

```{r}
pairwise.wilcox.test(data_50_runs$Score, data_50_runs$Search_method)
```

All the differences are strongly significant, with $p<2^{-16}$ in each case. Note, however, that while `Random` is better than `HC_zero`, the median is about the same for the two, so we wouldn't expect _huge_ differences between then. Certainly the improvements in performance that we see with `HC_penalty` are much more interesting.

## How do things change by problem? Max evals?

We saw earlier that there was some difference between the 20 and 200 item problems, because all the negative final results from `HC_penalty` were on 200 item problems. This plot shows the performance on all six problems, regardless of search method used:

```{r}
plot(data_50_runs$Score ~ data_50_runs$Problem,
     xlab="Searcher", ylab="Score")
```

There are clearly differences. Some, such as the much higher values on the rightmost boxplot, are likely at least partly because of differences in the maximum possible values of the problems. Others seem to be more about the difficulty of the problems; the second problem (`knapPI_11_200_1000_4`) has a lower median than several other problems despite having an apparently higher possible value, suggesting that it's harder than those problems (at least for these searchers).

The following plot shows the performance broken out by essentially _all_ our independent variables: Searcher, problem, and `max-tries`.

```{r warning=FALSE}
library("ggplot2")

ggplot(data_50_runs, 
       aes(x=factor(Max_evals), y=Score, group=Max_evals)) + 
  geom_boxplot() + facet_grid(Search_method ~ Problem)
```

Reading this horizontally shows differences in the problems with, for example, `knapPI_16_200_1000_4` clearly having much higher values (at least for `HC_penalty`) than any of the other problems. Reading the columns vertically shows differences across searchers for a specific problem; it's clear for example that whatever advantage `HC_penalty` has is *much* stronger on the 200 item problems, where the other two searchers never get above zero.

This also suggests that using 10,000 tries instead of 1,000 often didn't change things much. There are exceptions (e.g., `knapPI_16_2000_1000_4` on `HC_penalty` again), but typically the medians are quite close. This suggests that we might stick to 1,000 tries in future _initial_ explorations, and only switch to larger number of tries when we've identified which searchers, etc., we're especially interested in.

## Recursive partitioning

The results in the previous plot separating things by problem, searcher, and `max-tries` suggests that the interactions of the independent variables is somewhat complex, so I used `rpart` to try to understand the relative importance of the many differences.

```{r}
library("rpart")
library("rpart.plot")

rp <- rpart(Score ~ Search_method + Problem + Max_evals, data=data_50_runs)
rp

rpart.plot(rp, type=3, extra=100)
```

This indicates that despite the various differences between problems and different values of maximum evaluations, the choice of search searcher is the most important first-order difference, splitting on `HC_penalty` (on the right) vs. the other two searchers. After that split, though, the problems were the next most important factor along both branches. Focussing on the more interesting searcher (`HC_penalty`), `knapPI_16_200_1000_4` was "different" than the others, which isn't surprising given the substantially higher maximum values found on that problem than on the other problems. Once `rpart` is focusing on that particular problem, it also highlights the substantial difference between the 1,000 and 10,000 maximum evaluation runs.

# Conclusions

Based on these runs, it's clear that at least for these six problems `HC_penalty` is consistently as good or, in some cases, substantially better than the other two searchers tried here. This suggests that having a gradient to act on in the "illegal" part of the search space is a significant advantage on these problems.

Have more evaluations does sometimes help, and occassionally quite a bit, but it often doesn't make a substantial difference, especially on the easier problems. So I might consider starting with just 1,000 evaluations in future explorations, saving the higher number of evaluations for when I've narrowed down the pool of search tools I really want to explore more deeply. (That would also be a good time to include some test problems with more items.)

Laslty, the facetted plot and the `rpart` analysis make it clear that I _really_ should normalize my data by dividing all my scores by the highest score found for a given problem. That would reduce effects caused by disproportionate maximum values for problems like `knapPI_16_200_1000_4`, and allow tools like `rpart` to focus on differences caused by the choice of searchers or maximum evaluations.
